Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-05-25 Paper/paper.tex
text on nsbhs

\red{
Text about CBCs. How was each type observed so far?
}
\prayush{
Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf,LIGOVirgo2016a}.
% 
Compact binary merger rate estimates suggest
a GW detection rate of approximately a few tens of binary black holes
(BBH) every year~\cite{Abadie:2010cfa}. The actual masses of
astrophysical black holes are uncertain, but observations and
population synthesis studies suggest that BHs formed from stellar
core-collapse can have masses up to and higher than
$34M_\odot$~\cite{2008ApJ...678L..17S,2010ApJ...714.1217B}.  Also,
recent measurements using continuum fitting and X-ray reflection
fitting suggest that black holes can have high spin, with the BH
angular momentum in dimensionless units exceeding
$0.8$~\cite{2011ApJ...742...85G,2012MNRAS.424..217F,Gou:2013dna,
  2009ApJ...697..900M,McClintock:2006xd,Miller:2009cw,McClintock:2013vwa}.
Therefore the observations of GWs emitted by spinning BBHs will allow us to
understand the spin-spin and spin-orbit dynamics of the two-body
system, apart from allowing us to test strong-field dynamics of
General Relativity.
}
%

\red{
Text describing BHNS disruption and accretion dynamics.
}
An important sub-class of these
are NSBH binaries, of which we expect $\mathcal{O}(10)$ GW observations
every year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
%

\red{
Text focusing on tides
}
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.

The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible.
% 
\red{
Text about disruption of NSBHs, MOVE EARLIER.
}
\prayush{
The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.
}

@@

A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about binary neutron stars (BNS)~\cite{
Lee1999a,Lee1999b,Lee2000,oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,
stergioulas:11,East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. 
Studies~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a} have shown that
with $20-30$ BNS observations, aLIGO would begin to distinguish between hard,
moderate and soft candidate equations of state. These studies rely on the PN 
description of BNS dynamics, which is reasonable since BNSs merge at very high
frequencies and most of their signal power resides in inspiral.
% 
On the other hand, NSBHs merge at relatively lower frequencies than BNS, 
and their disruption during late-inspiral occurs at frequencies where aLIGO
has relatively more resolving power. This, combined with a shorter inspiral,
means that the merger portion of NSBH signals is relatively more valuable for
measuring $\lambdans$. Some past NSBH studies aimed at measuring tidal effects,
however, have used PN inspiral-only waveforms~\cite{Maselli:2013rza}. In doing
so, however, (i) they ignore the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms could easily dominate over the tidal terms themselves, since there 
are unknown point-particle terms at orders {\it lower} than that at which tidal
effects enter~\cite{Yagi:2014}.
\red{
Make clear the distinction between Fischer matrix studies and Bayesian ones. 
Explain the differences. this is important as it sets the stage for for what is
new in this paper.
} 
Other studies that
use complete numerical simulations or derived models either use Fisher matrix 
to estimate $\lambdans$ measurement error~\cite{Lackey2011,Lackey:2013axa},
which is not reliable at realistic signal-to-noise ratios 
(SNR)~\cite{Vallisneri:2007ev}, or are limited in the binary parameter space
they sample~\cite{Foucart:2013psa}.

>>>>

% \red{
% Text about CBCs. How was each type observed so far?
% }
Coalescing compact binaries (both monolithic and mixed) of stellar-mass black
holes (BH) and neutron stars (NS) are the primary target sources for the second
generation GW detectors~\cite{Timmes:1995kp,Fryer:1999mi,RevModPhys.74.1015,
2010ApJ...714.1217B,2010ApJ...715L.138B,Dominik:2014yma,Belczynski:2006zi,
2012ApJ...749...91F,
Wex:1998wt,1991ApJ...379L..17N,Mandel:2015spa,Abbott:2016nhf}.
A monolithic binary of black holes was recently observed by 
aLIGO~\cite{LIGOVirgo2016a}. Previously, stellar-mass black holes had only 
been seen in binaries with stellar companion (inferred from the observed spectrum
of electromagnetic waves emitted by the companion)~\cite{Lewin2010,
Remillard:2006fc,Fragos:2010tm}.
Neutron stars, on the other hand, have had numerous sightings. Thousands have
electromagnetically emitting neutron stars, or pulsars, have been 
documented~\cite{Manchester:2004bp},
in varied situations: as radio pulsars~\cite{Lattimer:2012nd,Manchester:2004bp},
in binary systems with stellar companions~\cite{1971ApJ...169L..23M,
Bond:2002eh,Lattimer:2012nd,Manchester:2004bp},
and in monolithic neutron star binaries~\cite{Hulse:1975uf,Taylor:1982wi,
Weisberg:2010zz,Lattimer:2012nd,Manchester:2004bp}.
Mixed binaries of black holes and neutron stars, is an astrophysically
interesting class of systems~\cite{Wex:1998wt,
1991ApJ...379L..17N,Janka1999,Fryer:2015jpa}, that is yet undetected.
We expect to observe $\mathcal{O}(10)$ mixed binaries per year with
aLIGO~\cite{Abadie:2010cf}.

% \prayush{
% The actual masses of
% astrophysical black holes are uncertain, but observations and
% population synthesis studies suggest that BHs formed from stellar
% core-collapse can have masses up to and higher than
% $34M_\odot$~\cite{2008ApJ...678L..17S,2010ApJ...714.1217B}.  Also,
% recent measurements using continuum fitting and X-ray reflection
% fitting suggest that black holes can have high spin, with the BH
% angular momentum in dimensionless units exceeding
% $0.8$~\cite{2011ApJ...742...85G,2012MNRAS.424..217F,Gou:2013dna,
%   2009ApJ...697..900M,McClintock:2006xd,Miller:2009cw,McClintock:2013vwa}.
% }
% %

% \red{
% Text describing BHNS disruption and accretion dynamics.
% }

% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
%

NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (SGRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Depending on their equation of state (EoS), NSs can get disrupted by
the tidal field of their companion BHs. Once disrupted, most of the NS
material falls into the hole over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest partly getting ejected as unbound material
% (kilonovae, r-process material),
and partly forming an accretion disk around the BH.
% 
This short lived ($0.1-1s$) disk-BH system is hypothesized to drive SGRBs
through the production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
% 
However, whether or not such a system forms depends also on the nature of
the BH. For e.g., massive BHs (with $\mbh\gtrsim 10M_\odot$), as well as BHs with
large retrograde spins, tend to swallow the NS whole without forming a
disk~\cite{Foucart:2013psa}. 
% 
On the other hand, {\it low-mass} BHs with $\mbh\in[3M_\odot, 7M_\odot]$, 
can disrupt their companion NSs much before merger, forming long-sustained disks
that are required to sustain 
SGRBs~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,Kawaguchi:2015}.
% 
% Such systems additionally leave a strong imprint on the emitted GW signal.
A {\it coincident} detection of both GWs and gamma-rays from an NSBH merger will
provide us with a uniquely opportunity to confirm this hypothesized link between
NSBH mergers and GRBs~\cite{Abbott:2016wya}.

\red{
Text focusing on tides
}

Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment of the star is proportional to the tidal field of its companion,
% through the relation $Q_{ij} = -\lambda\mathcal{E}_{ij}$,
with $\lambdans$ the EoS-dependent proportionality constant that is also the
tidal deformability parameter, related to the NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambdans:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
% 
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.

@@

GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs.

The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible.
% 
A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about binary neutron stars (BNS)~\cite{
Lee1999a,Lee1999b,Lee2000,oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,
stergioulas:11,East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. 
Studies~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a} have shown that
with $20-30$ BNS observations, aLIGO would begin to distinguish between hard,
moderate and soft candidate equations of state. These studies rely on the PN 
description of BNS dynamics, which is reasonable since BNSs merge at very high
frequencies and most of their signal power resides in inspiral.
% 
On the other hand, NSBHs merge at relatively lower frequencies than BNS, 
and their disruption during late-inspiral occurs at frequencies where aLIGO
has relatively more resolving power. This, combined with a shorter inspiral,
means that the merger portion of NSBH signals is relatively more valuable for
measuring $\lambdans$. Some past NSBH studies aimed at measuring tidal effects,
however, have used PN inspiral-only waveforms~\cite{Maselli:2013rza}. In doing
so, however, (i) they ignore the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms could easily dominate over the tidal terms themselves, since there 
are unknown point-particle terms at orders {\it lower} than that at which tidal
effects enter~\cite{Yagi:2014}.
\red{
Make clear the distinction between Fischer matrix studies and Bayesian ones. 
Explain the differences. this is important as it sets the stage for for what is
new in this paper.
} 
Other studies that
use complete numerical simulations or derived models either use Fisher matrix 
to estimate $\lambdans$ measurement error~\cite{Lackey2011,Lackey:2013axa},
which is not reliable at realistic signal-to-noise ratios 
(SNR)~\cite{Vallisneri:2007ev}, or are limited in the binary parameter space
they sample~\cite{Foucart:2013psa}.
